---
title: "Quiz 3"
date: "2/15/2021"
output: 
  html_document:
    number_sections: true
---
Name: __________________________
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r, echo=FALSE}
df <- data.frame(matrix(1:100, nr=10))

```


```{r}
df
```

1.  Consider the dataframe above called `df`.  What would running this code return `sum(df[7,7:10])`

```{r, echo=FALSE, eval=FALSE}
df <- data.frame(matrix(1:100, nr=10))
sum(df[7,7:10])
```

2. How can you tell if an object in R is a dataframe?


3. How could you create the dataframe above called `df`?


4. What code would return the average of row 2 of `df`?


5. Consider `mtcars` dataset that comes preloaded with R that looks like this:
```{r out.width="50%"}
head(mtcars)
```

6. Why do I get this error when I run the code below: `Error in plot(hp, mpg) : object 'hp' not found`?

```{r eval=FALSE}
plot(hp,mpg)
```

`Error in plot(hp, mpg) : object 'hp' not found`

7. What code would generate this plot?

```{r echo=FALSE, out.width="50%"}
attach(mtcars)
plot(hp,mpg)
```

Bonus: What is a topic that you find confusing at this point in class? 