---
title: "In-class fun with dates, times, data viz and wrangling"
author: "Tobin Turner"
date: "2/10/2021"
output: 
  html_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-packages, message=FALSE, warning=FALSE}
library("tidyverse")
library("lubridate")
library("nycflights13")
```


# Let's make sure we like and understand dates and times:

## Point of these questions?

-  Does every year have 365 days?
-  Does every day have 24 hours?
-  Does every minute have 60 seconds?

## Do these code chunks make sense:


```{r}
today()
now()
```


```{r}
flights %>% 
  select(year, month, day, hour, minute) %>% 
  mutate(departure = make_datetime(year, month, day, hour, minute))
```

```{r}
make_datetime_100 <- function(year, month, day, time) {
  make_datetime(year, month, day, time %/% 100, time %% 100)
}

flights_dt <- flights %>% 
  filter(!is.na(dep_time), !is.na(arr_time)) %>% 
  mutate(
    dep_time = make_datetime_100(year, month, day, dep_time),
    arr_time = make_datetime_100(year, month, day, arr_time),
    sched_dep_time = make_datetime_100(year, month, day, sched_dep_time),
    sched_arr_time = make_datetime_100(year, month, day, sched_arr_time)
  ) %>% 
  select(origin, dest, ends_with("delay"), ends_with("time"))

flights_dt
```

### Hadley says: "With this data, I can visualise the distribution of departure times across the year"
```{r}
flights_dt %>% 
  ggplot(aes(dep_time)) + 
  geom_freqpoly(binwidth = 86400) # 86400 seconds = 1 day
```

## The following code from the chapter is used.  Does it make sense?

```{r}
make_datetime_100 <- function(year, month, day, time) {
  make_datetime(year, month, day, time %/% 100, time %% 100)
}

flights_dt <- flights %>%
  filter(!is.na(dep_time), !is.na(arr_time)) %>%
  mutate(
    dep_time = make_datetime_100(year, month, day, dep_time),
    arr_time = make_datetime_100(year, month, day, arr_time),
    sched_dep_time = make_datetime_100(year, month, day, sched_dep_time),
    sched_arr_time = make_datetime_100(year, month, day, sched_arr_time)
  ) %>%
  select(origin, dest, ends_with("delay"), ends_with("time"))
```


```{r}
sched_dep <- flights_dt %>%
  mutate(minute = minute(sched_dep_time)) %>%
  group_by(minute) %>%
  summarise(
    avg_delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )
```

## What do these do: `floor_date()`, `round_date()`, and `ceiling_date()`?

## Try these

1. Look at the number of cancelled flights per day. Is there a pattern? Is the proportion of cancelled flights related to the average delay?

```{r, echo=FALSE, warning=F, message=FALSE}
# 5.6.7 Exercises # 4
cancelled_per_day <- 
  flights %>%
  mutate(cancelled = (is.na(arr_delay) | is.na(dep_delay))) %>%
  group_by(year, month, day) %>%
  summarise(
    cancelled_num = sum(cancelled),
    flights_num = n(),
  )
```


1. Which carrier has the worst delays? Challenge: can you disentangle the effects of bad airports vs. bad carriers? Why/why not? (Hint: think about `flights %>% group_by(carrier, dest) %>% summarise(n())`)

```{r}

```


1. Use a good old visual aid to illustrate if/how how the average delay time changes over the course of a day. Your graph should answer, do flights get more late as the day goes on?

```{r, eval=FALSE, echo=FALSE, message=FALSE}
#16.3.4
flights_dt %>%
  mutate(sched_dep_hour = hour(sched_dep_time)) %>%
  group_by(sched_dep_hour) %>%
  summarise(dep_delay = mean(dep_delay)) %>%
  ggplot(aes(y = dep_delay, x = sched_dep_hour)) +
  geom_point() +
  geom_smooth()
```


2. On what day of the week should you leave if you want to minimize the chance of a delay? How do you know? Use a bar chart to show the average delay for each day of the week.

```{r, eval=FALSE, echo=FALSE, message=FALSE}
# 16.3.5
flights_dt %>%
  mutate(dow = wday(sched_dep_time)) %>%
  group_by(dow) %>%
  summarise(
    dep_delay = mean(dep_delay),
    arr_delay = mean(arr_delay, na.rm = TRUE)
  ) %>%
  print(n = Inf)

flights_dt %>%
  mutate(wday = wday(dep_time, label = TRUE)) %>% 
  group_by(wday) %>% 
  summarize(ave_dep_delay = mean(dep_delay, na.rm = TRUE)) %>% 
  ggplot(aes(x = wday, y = ave_dep_delay)) + 
  geom_bar(stat = "identity")

flights_dt %>% 
  mutate(wday = wday(dep_time, label = TRUE)) %>% 
  group_by(wday) %>% 
  summarize(ave_arr_delay = mean(arr_delay, na.rm = TRUE)) %>% 
  ggplot(aes(x = wday, y = ave_arr_delay)) + 
  geom_bar(stat = "identity")
```


3. Are early arrivals of flights caused more by scheduled flights that leave early or that make up time in the air?

```{r echo=FALSE, eval=FALSE, message=FALSE}
# 16.3.7

flights_dt %>% 
  mutate(minute = minute(dep_time),
         early = dep_delay < 0) %>% 
  group_by(minute) %>% 
  summarise(
    early = mean(early, na.rm = TRUE),
    n = n()) %>% 
  ggplot(aes(minute, early)) +
    geom_line()
```

