---
title: "in class times, viz and wrangling"
author: "Tobin Turner"
date: "2/9/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r load-packages, message=FALSE}
library("tidyverse")
library("lubridate")
library("nycflights13")
```



```{r, echo=FALSE, message=FALSE}
make_datetime_100 <- function(year, month, day, time) {
  make_datetime(year, month, day, time %/% 100, time %% 100)
}

flights_dt <- flights %>%
  filter(!is.na(dep_time), !is.na(arr_time)) %>%
  mutate(
    dep_time = make_datetime_100(year, month, day, dep_time),
    arr_time = make_datetime_100(year, month, day, arr_time),
    sched_dep_time = make_datetime_100(year, month, day, sched_dep_time),
    sched_arr_time = make_datetime_100(year, month, day, sched_arr_time)
  ) %>%
  select(origin, dest, ends_with("delay"), ends_with("time"))
```

The following code from the chapter is used
```{r}
sched_dep <- flights_dt %>%
  mutate(minute = minute(sched_dep_time)) %>%
  group_by(minute) %>%
  summarise(
    avg_delay = mean(arr_delay, na.rm = TRUE),
    n = n()
  )
```


In the previous code, the difference between rounded and un-rounded dates provides the within-period time.

## Try these

1. Use the graph to show how the average delay time changes over the course of a day. Your graph should answer, do flights get more late as the day goes on?

```{r, echo=FALSE, message=FALSE}
#16.3.4
flights_dt %>%
  mutate(sched_dep_hour = hour(sched_dep_time)) %>%
  group_by(sched_dep_hour) %>%
  summarise(dep_delay = mean(dep_delay)) %>%
  ggplot(aes(y = dep_delay, x = sched_dep_hour)) +
  geom_point() +
  geom_smooth()
```


2. On what day of the week should you leave if you want to minimize the chance of a delay? How do you know? Use a bar chart to show the average delay for each day of the week.

```{r, echo=FALSE, message=FALSE}
# 16.3.5
flights_dt %>%
  mutate(dow = wday(sched_dep_time)) %>%
  group_by(dow) %>%
  summarise(
    dep_delay = mean(dep_delay),
    arr_delay = mean(arr_delay, na.rm = TRUE)
  ) %>%
  print(n = Inf)

flights_dt %>%
  mutate(wday = wday(dep_time, label = TRUE)) %>% 
  group_by(wday) %>% 
  summarize(ave_dep_delay = mean(dep_delay, na.rm = TRUE)) %>% 
  ggplot(aes(x = wday, y = ave_dep_delay)) + 
  geom_bar(stat = "identity")

flights_dt %>% 
  mutate(wday = wday(dep_time, label = TRUE)) %>% 
  group_by(wday) %>% 
  summarize(ave_arr_delay = mean(arr_delay, na.rm = TRUE)) %>% 
  ggplot(aes(x = wday, y = ave_arr_delay)) + 
  geom_bar(stat = "identity")
```


3. Are early arrivals of flights caused more by scheduled flights that leave early or that make up time in the air?

```{r, echo=FALSE, message=FALSE}
# 16.3.7

flights_dt %>% 
  mutate(minute = minute(dep_time),
         early = dep_delay < 0) %>% 
  group_by(minute) %>% 
  summarise(
    early = mean(early, na.rm = TRUE),
    n = n()) %>% 
  ggplot(aes(minute, early)) +
    geom_line()
```

