---
author: "Tobin Turner"
date: "2/17/2021"
output: 
  html_document:
    number_sections: true
---

# Linear Regression Worksheet

## `women` dataset

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Load and use help to understand the `women` dataset to answer the following questions:

### Create this figure to **guesstimate** the weight of someone who is 110 inches tall using a graph.

```{r echo=FALSE, out.width="30%", fig.align="center"}
attach(women)
m1 <- lm(weight~height,data = women)
plot(height,weight)
abline(m1)
```

### Calculate **precisely** the estimated weight of someone who is 110 inches tall.

## Class Data 

### Use the shoe size and height data from class to create a plot showing the relationship between height and shoe size.

### Create a best fit line on this graph.

### Predict the shoe size for a person 125 inches tall.

### Predict the height of a person with a shoe size of 2.55

### Create a graph that hours studied and GPA.
 
### Create a best fit line between hours studied and GPA. Predict GPA of a person who studies 512 hours per week.

### Create a best fit line between height and GPA. Does this look good? Predict GPA of a person who stands 90 inches tall.