---
title: "More and better data wrangling"
author: "BADM 371"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

> Check Out: https://dplyr.tidyverse.org/reference/

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(dplyr)
options(max.print = 7)
```

```{r, warning=FALSE, message=FALSE, out.width = "50%"}
# tally() is short-hand for summarise()
mtcars %>% tally()
mtcars %>% group_by(cyl) %>% tally()
# count() is a short-hand for group_by() + tally()
mtcars %>% count(cyl)
# Note that if the data is already grouped, count() adds
# an additional group that is removed afterwards
mtcars %>% group_by(gear) %>% count(carb)
# add_tally() is short-hand for mutate()
mtcars %>% add_tally()
# add_count() is a short-hand for group_by() + add_tally()
mtcars %>% add_count(cyl)
# count() and tally() are designed so that you can call them repeatedly, each time rolling up a level of detail
species <-  starwars %>%
 count(species, homeworld, sort = TRUE)
species
species %>% count(species, sort = TRUE)
# Change the name of the newly created column:
starwars %>%
 count(species, homeworld, sort = TRUE, name = "n_species_by_homeworld")
species %>%
 count(species, sort = TRUE, name = "n_species")
# add_count() is useful for groupwise filtering
# e.g.: show details for species that have a single member
starwars %>%
  add_count(species) %>%
  filter(n == 1)
rename(iris, petal_length = Petal.Length)
rename_with(iris, toupper)

rename_with(iris, toupper, starts_with("Petal"))
```

# Excercises

Using the `nycflights13` package and the `flights` dataset, answer these questions.  Some answers are given in square brackets for you to check your answers.

1. Which destination had the most flights with departure delays of greater than 60 minutes in this data set? [ORD with 1480]
2. What was the longest arrival delay in this dataset? [1272]
3. Which carrier in September had the most late departing flights? [UA with 1559]
4. Create a variable called total.annoyance which is arrival delay plus the departure delay for each flight.
5. Find all flights that Were operated by United, American, or Delta.
6. How many flights have a missing dep_time? 
7. Which carrier with more than 10 flights in September had greatest % late departing flights? 



