---
title: 'R: Self Learn Basics Part Deaux'
author: "Tobin Turner"
date: "2/10/2021"
output: 
  html_document:
    number_sections: true
---
# Let's learn us some MORE R

## Attaching to objects
R includes a number of datasets that it is convenient to use for examples. You can get a description
of what's available by typing

```{r}
data()
```

To access any of these datasets, you then type data(dataset) where dataset is the name of the
dataset you wish to access. For example,

```{r}
data(trees)
trees[1:5,]
```



gives us the first 5 rows of these data, and we can now see that the columns represent measurements
of girth, height and volume of trees (actually cherry trees: see help(trees)) respectively. Now, if we want to work on the columns of these data, we can use the subscripting technique
explained above: for example, trees[,2] gives all of the heights. This is a bit tedious however, and
it would be easier if we could refer to the heights more explicitly. We can achieve this by attaching
to the trees dataset:

```{r}
attach(trees)
```


Effectively, this makes the contents of trees a directory, and if we type the name of an object, R
will look inside this directory to find it. Since Height is the name of one of the columns of trees, R now recognises this object when we type the name. Hence, for example,

```{r}
mean(Height)
mean(trees[,2])
```


are synonymous, while it is easier to remember exactly what calculation is being performed by
the first of these expressions. In actual fact, trees is an object called a dataframe, essentially a
matrix with named columns (though a dataframe, unlike a matrix, may also include non-numerical
variables, such as character names). Because of this, there is another equivalent syntax to extract,
for example, the vector of heights:

```{r}
trees$Height
```


which can also be used without having first attached to the dataset.

## Exercises

1. Attach to the dataset quakes and produce a statistical summary of the variables depth and
mag.
2. Attach to the dataset mtcars and find the mean weight and mean fuel consumption for
vehicles in the dataset (type help(mtcars) for a description of the variables available).

## The apply function

It is possible to write loops in R, but they are best avoided whenever possible. A common situation
is where we want to apply the same function to every row or column of a matrix. For example,
we may want to find the mean value of each variable in the trees dataset. Obviously, we could
operate on each column separately but this can be tedious, especially if there are many columns.
The function apply simplifies things. It is easiest understood by example:

```{r}
apply(trees,2,mean)
```

has the effect of calculating the mean of each column (dimension 2) of trees. We'd have used a 1
instead of a 2 if we wanted the mean of every row.
Any function can be applied in this way, though if optional arguments to the function are
required these need to be specied as well - see help(apply) for further details.

## Exercises

1. Repeat the analyses of the datasets quakes and mtcars using the function apply to simplify
the calculations.
2. If
y =

```{r, echo =FALSE}
y <- c(1,4,1,0,2,1)
y <- matrix(y,nr=2,byrow = T)
y
```

what is the result of apply(y[,2:3],1,mean)? Check your answer in R.

## Graphics
R has many facilities for producing high quality graphics. A useful facility before beginning is to
divide a page into smaller pieces so that more than one figure can be displayed. For example:

```{r eval=FALSE}
par(mfrow=c(2,2))
```

creates a window of graphics with 2 rows and 2 columns. With this choice the windows are filled
up row-wise. Use mfcol instead of mfrow to fill up column-wise. The function par is a general
function for setting graphical parameters. There are many options: see help(par).

So, for example

```{r, out.width = '40%', fig.align = "center", fig.cap = "Figure 1: Tree heights and volumes"}
par(mfrow=c(2,2))
hist(Height)
boxplot(Height)
hist(Volume)
boxplot(Volume)
par(mfrow=c(1,1))
```

produces Figure 1. Note the final use of par to return the graphics window to standard size.
```{r fig.cap = "Figure 2: Scatterplot of tree heights and volumes", out.width = '40%', fig.align = "center"}
plot(Height,Volume)
```

We can also plot one variable against another using the function plot:

```{r fig.cap = "Figure 3: Scatterplot matrix for tree data", out.width = '40%', fig.align = "center"}
pairs(trees)
```


See Figure 3. Like many other functions plot is object-specic: its behaviour depends on the
object to which it is applied. For example, if the object is a matrix, plot is identical to pairs: try
plot(trees). For some other possibilities try:

```{r eval=FALSE}
data(nhtemp)
plot(nhtemp)

data(faithful)
plot(faithful)

data(HairEyeColor)
plot(HairEyeColor)
```


There are also many optional arguments in most plotting functions that can be used to control
colours, plotting characters, axis labels, titles etc. The functions points and lines are useful for
adding points and lines respectively to a current graph. The function abline is useful for adding
a line with specified intercept and slope.
To print a graph, point the cursor over the graphics window and press the right button on the
mouse. This should open up a menu which includes `print' as an option. You also have the option
to save the figure in various formats, for example as a postscript file, for storage and later use.


## Exercises

1. Use
> x<-rnorm(100)

or something similar, to generate some data. Produce a figure showing a histogram and
boxplot of the data. Modify the axis names and title of the plot in an appropriate way.

2. Type the following
```{r eval=FALSE}
x<- (-10):10
n<-length(x)
y<-rnorm(n,x,4)
plot(x,y)
abline(0,1)
```



Try to understand the effect of each command and the graph that is produced.

3. Type the following:

```{r eval=FALSE}
data(nhtemp)
plot(nhtemp)
```

This produces a time series plot of measurements of annual mean temperatures in New Hampshire, U.S.A.

4. The previous example illustrated that plot acts differently on objects of different types - the
object nhtemp has the special class time series. More generally, we may have the data of
yearly observations in a vector, but need to build the time series plot for ourselves. We can
mimic this situation by typing

`temp<-as.vector(nhtemp)`

which creates a vector temp that contains only the annual temperatures. We can produce
something similar to the time series plot by typing
`plot(1912:1971,temp)`
but this plots points rather than lines. To join the data via lines we would use
`plot(1912:1971,temp,type='l')`
instead. To get points and lines, use `type='b'` instead.
