---
title: 'R: Self Learn Basics'
author: "Tobin Turner"
date: "1/26/2021"
output: 
  html_document:
    number_sections: true
---
# Let's learn us some R

## Introduction

R is a software language for carrying out complicated (and simple) statistical analyses. It includes
routines for data summary and exploration, graphical presentation and data modelling. The aim of
this document is to provide you with a basic fluency in the language. It is suggested that you work
through this document at the computer, having started an R session. Type in all of the commands
that are printed, and check that you understand how they operate. Then try the simple exercises
at the end of each section.

When you work in R you create objects that are stored in the current workspace (sometimes
called image). Each object created remains in the image unless you explicitly delete it. At the end
of the session the workspace will be lost unless you save it. You can save the workspace at any
time by clicking on the disc icon at the top of the control panel.
Commands written in R are saved in memory throughout the session. You can scroll back to
previous commands typed by using the ‘up’ arrow key (and ‘down’ to scroll back again). You can
also ‘copy’ and ‘paste’ using standard windows editor techniques (for example, using the ‘copy’
and ‘paste’ dialog buttons). If at any point you want to save the transcript of your session, click
on ‘File’ and then ‘Save History’, which will enable you to save a copy of the commands you have
used for later use. As an alternative you might copy and paste commands manually into a notepad
editor or something similar.


## Objects and Arithmetic

R stores information and operates on objects. The simplest objects are scalars, vectors and matrices.
But there are many others: lists and dataframes for example. In advanced use of R it can also be
useful to define new types of object, specific for particular application. We will stick with just the
most commonly used objects here.
An important feature of R is that it will do different things on different types of objects. For
example, type:1
```{r}
4+6
```


So, R does scalar arithmetic returning the scalar value 10. (In actual fact, R returns a vector of
length 1 - hence the [1] denoting first element of the vector.
We can assign objects values for subsequent use. For example:
```{r}
x<-6
y<-4
z<-x+y
```


would do the same calculation as above, storing the result in an object called z. We can look at
the contents of the object by simply typing its name:
```{r}
z
```

At any time we can list the objects which we have created:
```{r}
ls()
```

Notice that ls is actually an object itself. Typing ls would result in a display of the contents of
this object, in this case, the commands of the function. The use of parentheses, ls(), ensures that
the function is executed and its result - in this case, a list of the objects in the directory - displayed.
More commonly a function will operate on an object, for example
```{r}
sqrt(16)
```

calculates the square root of 16. Objects can be removed from the current workspace with the rm
function:
```{r}
rm(x,y)
```

for example.
There are many standard functions available in R, and it is also possible to create new ones.
Vectors can be created in R in a number of ways. We can describe all of the elements:
```{r}
z<-c(5,9,1,0)
```

Note the use of the function c to concatenate or ‘glue together’ individual elements. This function
can be used much more widely, for example
```{r}
x<-c(5,9)
y<-c(1,0)
z<-c(x,y)
```


would lead to the same result by gluing together two vectors to create a single vector.
Sequences can be generated as follows:
```{r}
x<-1:10
```

while more general sequences can be generated using the seq command. For example:
```{r}
seq(1,9,by=2)
seq(8,20,length=6)
```

These examples illustrate that many functions in R have optional arguments, in this case, either
the step length or the total length of the sequence (it doesn’t make sense to use both). If you leave
out both of these options, R will make its own default choice, in this case assuming a step length
of 1. So, for example,
```{r}
seq(8,20,length=6)
x<-seq(1,10)
x
```

also generates a vector of integers from 1 to 10.
At this point it’s worth mentioning the help facility. If you don’t know how to use a function,
or don’t know what the options or default values are, type help(functionname) where functionname is the name of the function you are interested in. This will usually help and will often include
examples to make things even clearer.
Another useful function for building vectors is the rep command for repeating things. For
example
```{r}
rep(0,100)
rep(1:3,6)
rep(1:3,c(6,6,6))
```
which we could also simplify cleverly as
```{r}
rep(1:3,rep(6,3))
```

As explained above, R will often adapt to the objects it is asked to work on. For example:
```{r}
x<-c(6,8,9)
y<-c(1,2,4)
x+y
```
and
```{r}
x*y
```


showing that R uses componentwise arithmetic on vectors. R will also try to make sense if objects
are mixed. For example,
```{r}
x<-c(6,8,9)
x+2
```


though care should be taken to make sure that R is doing what you would like it to in these
circumstances.
Two particularly useful functions worth remembering are length which returns the length of a
vector (i.e. the number of elements it contains) and sum which calculates the sum of the elements
of a vector.

## Exercises
1. Define
```{r}
x<-c(4,2,6)
y<-c(1,0,-1)
```


2. Decide what the result will be of the following:
    + length(x)
    + sum(x)
    + sum(x^2)
    + x+y
    + x*y
    + x-2
    + x^2
3. Use R to check your answers.
4. Decide what the following sequences are and use R to check your answers:
    + 7:11
    + seq(2,9)
    + seq(4,10,by=2)
    + seq(3,30,length=10)
    + seq(6,-4,by=-2)
5. Determine what the result will be of the following R expressions, and then use R to check if	you are right:
    + rep(2,4)
    + rep(c(1,2),4)
    + rep(c(1,2),c(4,4))
    + rep(1:4,4)
    + rep(1:4,rep(3,4))
6. Use the rep function to define simply the following vectors in R.
    + 6,6,6,6,6,6
    + 5,8,5,8,5,8,5,8
    + 5,5,5,5,8,8,8,8

# R basics and workflows {#r-basics}

## Basics of working with R at the command line and RStudio goodies

Launch RStudio/R.

Notice the default panes:

* Console (entire left)
* Environment/History (tabbed in upper right)
* Files/Plots/Packages/Help (tabbed in lower right)

FYI: you can change the default location of the panes, among many other things: [Customizing RStudio][rstudio-customizing].

Go into the Console, where we interact with the live R process.

Make an assignment and then inspect the object you just created:

```{r}
x <- 3 * 4
x
```

All R statements where you create objects -- "assignments" -- have this form:
```{r eval = FALSE}
objectName <- value
```
and in my head I hear, e.g., "x gets 12".


You will make lots of assignments and the operator `<-` is a pain to type. Don't be lazy and use `=`, although it would work, because it will just sow confusion later. Instead, utilize RStudio's keyboard shortcut: Alt + - (the minus sign).

Notice that RStudio automagically surrounds `<-` with spaces, which demonstrates a useful code formatting practice. Code is miserable to read on a good day. Give your eyes a break and use spaces.

RStudio offers many handy [keyboard shortcuts][rstudio-key-shortcuts]. Also, Alt+Shift+K brings up a keyboard shortcut reference card.

Object names cannot start with a digit and cannot contain certain other characters such as a comma or a space. You will be wise to adopt a [convention for demarcating words][wiki-snake-case] in names.

```
i_use_snake_case
other.people.use.periods
evenOthersUseCamelCase
```

Make another assignment:
```{r}
this_is_a_really_long_name <- 2.5
```

To inspect this, try out RStudio's completion facility: type the first few characters, press TAB, add characters until you disambiguate, then press return.

Make another assignment:
```{r}
turner_rocks <- 2 ^ 3
```

Let's try to inspect:
```{r error=TRUE}
turnerrocks
Turner_rocks
turner_rocks
```

Implicit contract with the computer / scripting language: Computer will do tedious computation for you. In return, you will be completely precise in your instructions. Typos matter. Case matters. Get better at typing.

R has a mind-blowing collection of built-in functions that are accessed like so:

```{r eval = FALSE, tidy = FALSE}
functionName(arg1 = val1, arg2 = val2, and so on)
```

Let's try using `seq()` which makes regular sequences of numbers and, while we're at it, demo more helpful features of RStudio.

Type `se` and hit TAB. A pop up shows you possible completions. Specify `seq()` by typing more to disambiguate or using the up/down arrows to select. Notice the floating tool-tip-type help that pops up, reminding you of a function's arguments. If you want even more help, press F1 as directed to get the full documentation in the help tab of the lower right pane. Now open the parentheses and notice the automatic addition of the closing parenthesis and the placement of cursor in the middle.  Type the arguments `1, 10` and hit return. RStudio also exits the parenthetical expression for you.  IDEs are great.

```{r}
seq(1, 10)
```

The above also demonstrates something about how R resolves function arguments. You can always specify in `name = value` form. But if you do not, R attempts to resolve by position. So above, it is assumed that we want a sequence `from = 1` that goes `to = 10`. Since we didn't specify step size, the default value of `by` in the function definition is used, which ends up being 1 in this case. For functions I call often, I might use this resolve by position for the first
argument or maybe the first two. After that, I always use `name = value`.


Make this assignment and notice similar help with quotation marks.
```{r}
yo <- "hello world"
```

If you just make an assignment, you don't get to see the value, so then you're tempted to immediately inspect.

```{r}
y <- seq(1, 10)
y
```

This common action can be shortened by surrounding the assignment with parentheses, which causes assignment and "print to screen" to happen.

```{r}
(y <- seq(1, 10))
```

Not all functions have (or require) arguments:
```{r}
date()
```

Now look at your workspace -- in the upper right pane. The workspace is where user-defined objects accumulate. You can also get a listing of these objects with commands:

```{r}
objects()
ls()
```

If you want to remove the object named `y`, you can do this:

```{r}
rm(y)
```

To remove everything:

```{r}
rm(list = ls())
```

or click the broom in RStudio's Environment pane.

## Workspace and working directory

One day you will need to quit R, go do something else and return to your analysis later.

One day you will have multiple analyses going that use R and you want to keep them separate.

One day you will need to bring data from the outside world into R and send numerical results and figures from R back out into the world.

To handle these real life situations, you need to make two decisions:

* What about your analysis is "real", i.e. will you save it as your lasting record of what happened?
* Where does your analysis "live"?

### Workspace, `.RData`

As a beginning R user, it's OK to consider your workspace "real". _Very soon_, I urge you to evolve to the next level, where you consider your saved R scripts as "real". (In either case, of course the input data is very much real and requires preservation!) With the input data and the R code you used, you can reproduce
_everything_. You can make your analysis fancier. You can get to the bottom of puzzling results and discover and fix bugs in your code. You can reuse the code to conduct similar analyses in new projects. You can remake a figure with different aspect ratio or save is as TIFF instead of PDF. You are ready to take questions. You are ready for the future.

If you regard your workspace as "real" (saving and reloading all the time), if you need to redo analysis ... you're going to either redo a lot of typing (making mistakes all the way) or will have to mine your R history for the commands you used. Rather than [becoming an expert on managing the R history][rstudio-command-history], a better use of your time and psychic energy is to keep your "good" R code in a script for future reuse.

Because it can be useful sometimes, note the commands you've recently run appear in the History pane.

But you don't have to choose right now and the two strategies are not incompatible. Let's demo the save / reload the workspace approach.

Upon quitting R, you have to decide if you want to save your workspace, for potential restoration the next time you launch R. Depending on your set up, R or your IDE, e.g. RStudio, will probably prompt you to make this decision.

Quit R/RStudio, either from the menu, using a keyboard shortcut, or by typing `q()` in the Console. You'll get a prompt like this:

> Save workspace image to ~/.Rdata?

_Note where the workspace image is to be saved_ and then click "Save".

Using your favorite method, visit the directory where image was saved and verify there is a file named `.RData`. You will also see a file `.Rhistory`, holding the commands submitted in your recent session.

Restart RStudio. In the Console you will see a line like this:

```
[Workspace loaded from ~/.RData]
```


indicating that your workspace has been restored. Look in the Workspace pane and you'll see the same objects as before. In the History tab of the same pane, you should also see your command history. You're back in business. This way of starting and stopping analytical work will not serve you well for long but it's a start.

### Working directory

Any process running on your computer has a notion of its "working directory". In R, this is where R will look, by default, for files you ask it to load. It also where, by default, any files you write to disk will go. Chances are your current working directory is the directory we inspected above, i.e. the one where RStudio wanted to save the workspace.

You can explicitly check your working directory with:

```{r eval = FALSE}
getwd()
```

It is also displayed at the top of the RStudio console.

As a beginning R user, it's OK let your home directory or any other weird directory on your computer be R's working directory. _Very soon_, I urge you to evolve to the next level, where you organize your analytical projects into directories and, when working on project A, set R's working directory to the associated directory.

__Although I do not recommend it__, in case you're curious, you can set R's working directory at the command line like so:

```{r eval = FALSE}
setwd("~/myCoolProject")
```

__Although I do not recommend it__, you can also use RStudio's Files pane to navigate to a directory and then set it as working directory from the menu: *Session > Set Working Directory > To Files Pane Location*. (You'll see even more options there). Or within the Files pane, choose "More" and "Set As Working Directory".

But there's a better way. A way that also puts you on the path to managing your R work like an expert.

# Matrices, Summaries and Subscripting


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##  Summaries and Subscripting
Let's suppose we've collected some data from an experiment and stored them in an object x:

```{r}
x<-c(7.5,8.2,3.1,5.6,8.2,9.3,6.5,7.0,9.3,1.2,14.5,6.2)
```

Some simple summary statistics of these data can be produced:
```{r}
mean(x)
var(x)
sd(x)
summary(x)
```


which should all be self explanatory. 

It may be, however, that we subsequently learn that the first 6 data correspond to measurements made on one machine, and the second six on another machine.

This might suggest summarizing the two sets of data separately, so we would need to extract from
x the two relevant subvectors. This is achieved by subscripting:
```{r}
x[1:6]
x[7:12]
summary(x[1:6])
summary(x[7:12])
```

Other subsets can be created in the obvious way. For example:
```{r}
x[c(2,4,9)]
```

Negative integers can be used to exclude particular elements. For example
```{r}
x[-(1:6)]
```

has the same effect as 
```{r}
x[7:12]
```

### Exercises
1. If x<- c(5,9,2,3,4,6,7,0,8,12,2,9) decide what each of the following is and use R to
check your answers:
(a) x[2]
(b) x[2:4]
(c) x[c(2,3,6)]
(d) x[c(1:5,10:12)]
(e) x[-(10:12)]
2. The data y<-c(33,44,29,16,25,45,33,19,54,22,21,49,11,24,56) contain sales of milk
in gallons for 5 days in three different shops (the frst 3 values are for shops 1,2 and 3 on
Monday, etc.) Produce a statistical summary of the sales for each day of the week and also
for each shop.

### Matrices
Matrices can be created in R in a variety of ways. Perhaps the simplest is to create the columns
and then glue them together with the command cbind. For example,
Matrices can be created in R in a variety of ways. Perhaps the simplest is to create the columns
and then glue them together with the command cbind. For example,
```{r}
x<-c(5,7,9)
y<-c(6,3,4)
z<-cbind(x,y)
z
```


The dimension of a matrix can be checked with the dim command:
```{r}
dim(z)
```

[1] 3 2 i.e., three rows and two columns. There is a similar command, rbind, for building matrices by gluing rows together.  

The functions cbind and rbind can also be applied to matrices themselves (provided the dimensions match) to form larger matrices. For example,
```{r}
rbind(z,z)
```

Matrices can also be built by explicit construction via the function matrix. For example,
```{r}
z<-matrix(c(5,7,9,6,3,4),nrow=3)
```
results in a matrix z identical to z above. Notice that the dimension of the matrix is determined
by the size of the vector and the requirement that the number of rows is 3, as specified by the
argument nrow=3. As an alternative we could have specified the number of columns with the
argument ncol=2 (obviously, it is unnecessary to give both). Notice that the matrix is 'flled up'
column-wise. If instead you wish to fill up row-wise, add the option byrow=T. For example,
```{r}
z<-matrix(c(5,7,9,6,3,4),nr=3,byrow=T)
z
```
Notice that the argument nrow has been abbreviated to nr. Such abbreviations are always
possible for function arguments provided it induces no ambiguity - if in doubt always use the full
argument name.

As usual, R will try to interpret operations on matrices in a natural way. For example, with z
as above, and
```{r}
y<-matrix(c(1,3,0,9,5,-1),nrow=3,byrow=T)
y
```

we obtain
```{r}
y+z
```
and
```{r}
y*z
```

Other useful functions on matrices are to transpose a matrix:
```{r}
z
t(z)
```

As with vectors it is useful to be able to extract sub-components of matrices. In this case, we
may wish to pick out individual elements, rows or columns. As before, the [ ] notation is used to
subscript. The following examples should make things clear:
```{r}
z[1,1]
```

```{r}
z[c(2,3),2]
```

```{r}
z[,2]
```

> 
```{r}
z[1:2,]
```

So, in particular, it is necessary to specify which rows and columns are required, whilst omitting
the integer for either dimension implies that every element in that dimension is selected.

### Exercises


1. Create this matrix in R
```{r, echo = FALSE}
x <-matrix(c(1,7,8,11,-5,3,8,6,3,-9,0,11,14, 14, 14),nrow=3,byrow=T)
x
```
2. Create in R these matrices: 
```{r, echo = FALSE}
x <-matrix(c(1,7,8,11,5,9),nrow=3,byrow=T)
y <-matrix(c(6,8,2,1,1,-7),nrow=3,byrow=T)
x
y
```
3. Calculate the following and check your answers in R:
(a) 2*x
(b) x*x
(c) t(y)

```{r,echo = FALSE}
x <-matrix(c(1,7,8,11,5,9),nrow=3,byrow=T)
y <-matrix(c(6,8,2,1,1,-7),nrow=3,byrow=T)
2*x
x*x
t(y)
```

4. With x and y as above, calculate the effect of the following subscript operations and check
your answers in R.
(a) x[1,]
(b) x[2,]
(c) x[,2]
(d) y[1,2]
(e) y[,2:3]

