---
title: "Quiz Linear Regression ISLR Ch 3"
author: "Tobin Turner, adapted from ISLR Ch 3"
date: "3/3/2021"
output: 
  html_document:
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Quiz ISLR Chapter 3 Linear Regression

1. Define collinearity and give an example.

<br>

2. Define R^2^ and Adjusted R^2^

<br>

3. Explain least squares approach

<br>

4. Define p-value

<br>

5. In `m1`, below, which variables are significant predictors of Balance?  How do you know?

```{r load-packages, message=FALSE, warning=FALSE}
library("ISLR")
data(Credit)
attach(Credit)
head(Credit)
m1 <- lm(Balance ~ ., data = Credit)
summary(m1)
```

6. How "good" is the model created in `m1`?  How do you know and how could you improve this model?

<br>

<br>

7. Based on the model below, what would you predict the balance to be for an individual who is 55, has an income of $100,000, 13 years of education, is Caucasian and is a student? 
```{r}
m2 <- lm(Balance ~ Age + Income + Education + Ethnicity + Student, data = Credit)
summary(m2)
```

<br>

<br>

8. Interpret this model and its output, especially the coefficients `Income:Education   0.3149`:
```{r}
m3 <- lm(Balance ~ Income*Education, data = Credit)
summary(m3)
```

<br>

9. Something you found new and interesting in this chapter?